Discriminating the Influence of Correlated Factors \\from Multivariate Observations: the Back-to-Back Regression

### 2020-05-19 paper/appendix.tex
E -> S

     Consider the B2B model from equation \ref{eq:model} $$Y = (XE + N)F$$ with
     $N$ centered and full rank noise.

     If $F$ and $X$ are full-rank on $Img(E)$, then, the solution of B2B, $\hat
     H$ minimizes

@@

     $$E\hat H = \hat H$$
\end{theorem}
\begin{proof}

@@

 Since $\hat G$ is the least square estimator of $X$ from $Y$
 \begin{align*}
    \hat G = \arg \min_G \mathbb{E}[\left \| YG - X \right \|^2]
\end{align*}
Replacing $Y$ by its model definition $Y = (XE+N)F$, we have
 \begin{align*}
    \hat G &=   \arg \min_G \mathbb{E}[\left \| X - (XE + N)FG \right\|^2] =\arg \min_G \mathbb{E}[\left \| X - XEFG + NFG \right\|^2]
  \end{align*}
  Since $N$ is centered and independent of $X$, we have
  \begin{align}
    	  \hat G &=  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:Gdoublenorm}
\end{align}

In the same way, for $\hat H$, we have
\begin{align*}
    \hat H = \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
 \end{align*}
 a positive quantity which reaches a minimum (zero) for
 \begin{align}
    \hat H = EF \hat G
    \label{eq:Hdoublenom}
\end{align}

Let us now prove that $EF\hat G = F\hat G$.

Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger EF\hat G$, we
have $FZ = FF^\dagger EF \hat G$

Since $F$ is full rank on $Img(E)$, we have $FF^\dagger E =E$, and $FZ = EF\hat
G$

As $E$ is a binary diagonal matrix, it is an orthogonal projection and therefore
a contraction, thus
 $$ \| NEF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
 $$\left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$

But since $\hat G =  \arg \min_G \left \| X - XEFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XEFZ\right \| ^2  + \left \| NFZ\right \| ^2$$

Summarizing the above,
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2 \leq \| X - XEF\hat G \| ^2  + \| NF\hat G \| ^2$$
$$\left \| X - XEF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XEF\hat G \| ^2  + \| NEF\hat G \| ^2$$
$$\left \| NF\hat G\right \| ^2 =  \| NEF\hat G \| ^2$$

$N$ being full rank, this yields $EF\hat G = F\Hat G$.

Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = EFG$, we have
\begin{align*}
	\hat G &=  \arg \min_G  \left \| X - XEFG\right \| ^2  + \left \| NFG\right \| ^2 \\
	&=   \arg \min_G \left \| X - XEFG\right \| ^2  + \left \| NEFG\right \| ^2 \\
	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
	\label{eq:4}
\end{align*}

Finally, $E\hat H = E EF\hat G = EF\hat G = \hat H$, since $E$, a binary
diagonal matrix, is involutive. This completes the proof.
\end{proof}

@@

Equation \ref{eq:model} does not explicitly contain a measurement noise term.
Yet, in most experimental cases, the problem is best described as:
\begin{equation}
  Y = (XE+N)F+M
  \label{eq:model_noisy_measure}
\end{equation}
  with $M\in \mathbb{R}^{n \times d_y}$.

This equation is actually equivalent to Equation \ref{eq:model} given our
hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
$$Y = (XE+N)F+M = (XE+N+MF^{-1})F = (XE+N')F$$

Consequently, assuming that $F$ is full rank on $Img(XE)$, B2B yields the same
solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.

@@

\subsection{Recovering E}
\label{recovering}

In case of noise, B2B yields non binary $\hat E$. Three thresholding rules can
be used to binarize its values thus explicitly recover "causal" features.

First, given known signal-to-noise ratio, the threshold above which a feature
should considered to be "causal" can be derived analytically. Indeed, Equation
$\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
bounded:  % we assume that $F$ has full rank over $Img(E)$.

@@

>>>>

     Consider the B2B model from equation \ref{eq:model} $$Y = (XS + N)F$$ with
     $N$ centered and full rank noise.

     If $F$ and $X$ are full-rank on $Img(S)$, then, the solution of B2B, $\hat
     H$ minimizes

@@

     $$S\hat H = \hat H$$
\end{theorem}
\begin{proof}

@@

 Since $\hat G$ is the least square estimator of $X$ from $Y$
 \begin{align*}
    \hat G = \arg \min_G \mathbb{S}[\left \| YG - X \right \|^2]
\end{align*}
Replacing $Y$ by its model definition $Y = (XS+N)F$, we have
 \begin{align*}
    \hat G &=   \arg \min_G \mathbb{S}[\left \| X - (XS + N)FG \right\|^2] =\arg \min_G \mathbb{S}[\left \| X - XSFG + NFG \right\|^2]
  \end{align*}
  Since $N$ is centered and independent of $X$, we have
  \begin{align}
    	  \hat G &=  \arg \min_G \left \| X - XSFG\right\| ^2  + \left \| NFG\right \| ^2
     \label{eq:Gdoublenorm}
\end{align}

In the same way, for $\hat H$, we have
\begin{align*}
    \hat H = \arg \min_H \mathbb{S}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{S}[\| XH - (XS + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2] + \mathbb{S}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{S}[\| X(H - SF \hat G) \| ^2]
 \end{align*}
 a positive quantity which reaches a minimum (zero) for
 \begin{align}
    \hat H = SF \hat G
    \label{eq:Hdoublenom}
\end{align}

Let us now prove that $SF\hat G = F\hat G$.

Let $F^\dagger$ be the pseudo inverse of $F$, and $Z=F^\dagger SF\hat G$, we
have $FZ = FF^\dagger SF \hat G$

Since $F$ is full rank on $Img(S)$, we have $FF^\dagger S =S$, and $FZ = SF\hat
G$

As $S$ is a binary diagonal matrix, it is an orthogonal projection and therefore
a contraction, thus
 $$ \| NSF\hat G\|^2 \leq \| NF\hat G \|^2$$ and
 $$\left \| X - XSFZ\right \| ^2  + \left \| NFZ\right \| ^2 = \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NF\hat G \| ^2$$

But since $\hat G =  \arg \min_G \left \| X - XSFG\right\| ^2  + \left \| NFG\right \| ^2$, we also have
$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \left \| X - XSFZ\right \| ^2  + \left \| NFZ\right \| ^2$$

Summarizing the above,
$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2 \leq \| X - XSF\hat G \| ^2  + \| NF\hat G \| ^2$$
$$\left \| X - XSF\hat G\right\| ^2  + \left \| NF\hat G\right \| ^2 = \| X - XSF\hat G \| ^2  + \| NSF\hat G \| ^2$$
$$\left \| NF\hat G\right \| ^2 =  \| NSF\hat G \| ^2$$

$N$ being full rank, this yields $SF\hat G = F\Hat G$.

Replacing into $\eqref{eq:Gdoublenorm}$, and setting $H = SFG$, we have
\begin{align*}
	\hat G &=  \arg \min_G  \left \| X - XSFG\right \| ^2  + \left \| NFG\right \| ^2 \\
	&=   \arg \min_G \left \| X - XSFG\right \| ^2  + \left \| NSFG\right \| ^2 \\
	\hat H &=  \arg \min_H \left \| X - XH\right \| ^2  + \left \| NH\right \| ^2
	\label{eq:4}
\end{align*}

Finally, $S\hat H = S SF\hat G = SF\hat G = \hat H$, since $S$, a binary
diagonal matrix, is involutive. This completes the proof.
\end{proof}

@@

Equation \ref{eq:model} does not explicitly contain a measurement noise term.
Yet, in most experimental cases, the problem is best described as:
\begin{equation}
  Y = (XS+N)F+M
  \label{eq:model_noisy_measure}
\end{equation}
  with $M\in \mathbb{R}^{n \times d_y}$.

This equation is actually equivalent to Equation \ref{eq:model} given our
hypotheses. Indeed, we can rewrite $M = MF^{-1}F$ over $Img(F)$, which leads to:
$$Y = (XS+N)F+M = (XS+N+MF^{-1})F = (XS+N')F$$

Consequently, assuming that $F$ is full rank on $Img(XS)$, B2B yields the same
solutions to equations \ref{eq:model} and \ref{eq:model_noisy_measure}.

@@

\subsection{Recovering S}
\label{recovering}

In case of noise, B2B yields non binary $\hat S$. Three thresholding rules can
be used to binarize its values thus explicitly recover "causal" features.

First, given known signal-to-noise ratio, the threshold above which a feature
should considered to be "causal" can be derived analytically. Indeed, Equation
$\ref{eq:diagk}$ implies that the $k$ first diagonal elements of $\hat H$ are
bounded:  % we assume that $F$ has full rank over $Img(S)$.

@@

\begin{figure}
  \begin{center}
    \includegraphics[width=0.7\textwidth,
                     trim=0cm 0cm 0cm 0cm,
                     clip=True]{figures/fmri_decod.pdf}
    \includegraphics[width=0.7\textwidth,
                     trim=0cm 0cm 0cm 0cm,
                     clip=True]{figures/fmri_betas.pdf}

    \label{fig:fmri_supp}
  \end{center}
  \caption{Top. P-values across subjects derived from a Backward decoding of
  each feature (columns). Bottom. P-values across subjects derived from
  the $\beta$ coefficients of a Forward Model and the $\hat S$ of an unbiased
  B2B (i.e. B2B where $H$ is not regularized). Each cell displays the left
  hemisphere view laterally (top) or ventrally (bottom)}
\end{figure}

### 2020-05-19 paper/fmri_exps.tex
E -> S

We aim to identify the word features that cause a variation in brain responses. We consider four distinct but colinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is expected to primarily cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
expected to primarily cause a variation in the late evoked MEG responses (i.e.
from 400 ms) in the left frontal, temporal and parietal cortices
cortices \citep{kutas2011thirty,mccandliss2003visual}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a
noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
a conjunction, a determinant, a pronoun or a numeral), and was derived from
Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
not been thoroughly investigated with fMRI and MEG. While its causal contribution to
reading processes in the brain thus remains unclear, this lexical feature can
nonetheless be expected to present similar brain patterns to word frequency.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features
and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms
thus have a zero-mean and a unit-variance).

@@

In contrast, the $\hat E$ of an unbiased B2B (i.e. a B2B where $H$ is not
regularized) reaches higher levels of significance at each vertex
than the Forward $\beta$ \ref{fig:fmri_supp}. (Note that a direct
comparison between the Forward coefficients $\beta$ and those of B2B ($\hat E$)
is not feasible, as these two do not have the same units.)

>>>>

We aim to identify the word features that cause a variation in brain responses. We consider four distinct but collinear features.
%
First, 'Word Length' refers to the total number of letters. Word Length is expected to primarily cause a variation in the early evoked MEG responses (i.e. from 100 ms after stimulus onset) elicited by the retinotopically-tuned visual cortices (e.g. \citep{pegado2014timing}.).
%
Second, 'Word Frequency' indexes how frequently each word appears in Dutch and
was derived with the the Zipf logarithmic scale of \citep{van2014subtlex}
provided by the WordFreq package \citep{speerwordfreq}. Word Frequency is
expected to primarily cause a variation in the late evoked MEG responses (i.e.
from 400 ms) in the left frontal, temporal and parietal cortices
cortices \citep{kutas2011thirty,mccandliss2003visual}.
%
Third, 'Word Function' indicates whether each word is a content word (i.e. a
noun, a verb, an adjective or an adverb) or a function word (i.e. a preposition,
a conjunction, a determinant, a pronoun or a numeral), and was derived from
Spacy's part of speech tagger \citep{spacy2}. To our knowledge, this feature has
not been thoroughly investigated with fMRI and MEG. While its causal contribution to
reading processes in the brain thus remains unclear, this lexical feature can
nonetheless be expected to present similar brain patterns to word frequency.
%
Finally, to verify that B2B and other methods would not inadequately identify non-causal features, we added a dummy feature, constructed from a noisy combination of Word Length and Word Frequency:
$dummy = z(length) + z(frequency) + \mathcal{N}$, where $z$ normalizes features
and $\mathcal{N}$ is a random vector sampling Gaussian distribution (all terms
thus have a zero-mean and a unit-variance).

@@

In contrast, the $\hat S$ of an unbiased B2B (i.e. a B2B where $H$ is not
regularized) reaches higher levels of significance at each vertex
than the Forward $\beta$ \ref{fig:fmri_supp}. (Note that a direct
comparison between the Forward coefficients $\beta$ and those of B2B ($\hat S$)
is not feasible, as these two do not have the same units.)
